---
title: "Untitled"
format: html
editor: visual
---

# Data Visualization: Shiny App

The purpose of this document is to create a functioning Shiny application. The data will be a de-identified dataset from an anonymous database of clinical data.

## Packages:

```{r}
library(tidyverse)
library(ggplot2)
library(shiny)
```

## Data import

```{r}
data <- read.csv("C:\\Users\\domin\\Downloads\\Practicum_Cleaned_Data.csv")
```

Quick review for Project Description: Creating a Density plot for Baseline Seizure counts based on Insurance

```{r}
# Filter dataset to include only "Private" and "Medicaid/CHF"
filtered_insurance <- data %>%
  filter(insurance %in% c("Private", "Medicaid/CHF"))

# Create density plot
ggplot(filtered_insurance, aes(x = baseline_seizure, fill = insurance, color = insurance)) +
  geom_density(alpha = 0.3) +  # Transparency to see overlap
  scale_x_log10() +  # Log-scale if needed
  theme_minimal() +
  labs(title = "Density Plot of Count Variable by Insurance Type",
       x = "Intake Seizure Frequency",
       y = "Density",
       fill = "Insurance Type",
       color = "Insurance Type") +
  scale_fill_manual(values = c("Private" = "blue", "Medicaid/CHF" = "red")) +
  scale_color_manual(values = c("Private" = "blue", "Medicaid/CHF" = "red"))
```

Continuing this theme, let's create a box plot of baseline seizure:

```{r}
# Reshape data from wide to long format
long_data <- data %>%
  pivot_longer(cols = c(baseline_seizure, three_month_seizure, six_month_seizure, discharge_seizure), 
               names_to = "Variable", 
               values_to = "Value")

ggplot(long_data, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(alpha = 0.7) +  # Transparency to distinguish overlapping areas
  theme_minimal() +
  scale_y_log10() +
  labs(title = "Boxplot of Four Variables",
       x = "Timepoint",
       y = "Seizure Amount") +
  scale_fill_manual(values = c("baseline_seizure" = "blue", "three_month_seizure" = "red", "six_month_seizure" = "green", "discharge_seizure" = "purple"))


```

Now for another but only with patients that have depression:

```{r}


# Now let's try a combined plot with the entire population and only those with depression:

# Create a new column to label the two groups
long_data_combined <- long_data %>%
  mutate(Group = ifelse(depression_any == 1, "Depression", "No Depression"))

# Create the combined boxplot
ggplot(long_data_combined, aes(x = Variable, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.75)) +  # Dodge boxes for separation
  theme_minimal() +
  labs(title = "Comparison of All Patients vs. Patients w/ Depression Dx",
       x = "Timepoint",
       y = "Log of seizure frequency",
       fill = "Group") +
  scale_y_log10() +
  scale_fill_manual(values = c("No Depression" = "blue", "Depression" = "red"))
```
