---
title: ""
format: html
editor: visual
---

# R Animated Visualizations:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I want to use this document to practice using gganimate. The goal is to create animated visualizations to better describe visual data,

First, we need to set out libraries:

```{r}
library(gapminder)
library(tidyverse)
library(gganimate)
library(gifski)
library(png)
library(av)
```

I will be using some of the global health datasets from Gapminder: <https://www.gapminder.org/data/.>

My interest is the relationship between a country's C02 Emissions' and its GDP per Capita. I would like to model this over time. My hypothesis is that individual countries will have greater C02 emissions as their GDP increases.

```{r}

#To move foward with this, I downloaded a dataset for total C02 Emissions.

#Let's start by importing the dataset:


co2 <- read.csv("C:\\Users\\domin\\Downloads\\co2_pcap_cons.csv")

head(co2)

```

• Use Tidy Data functions to integrate the downloaded table(s) with the existing gapminder dataset in R. (3 points)

```{r}
# Let's look at the gaminder dataset first and see the levels we need to merge on.
gapminder

#Ok, the gapminder package is: A. in the long format, and B.starting at 1952 rather than 1800. Additionally, the downloaded frames have expected values past 2007. We'll need to remove the years from 1800 - 1952, remove the years after 2007, and then flip the data from wide to a long dataframe.

#Let's change the column names to only have the year

co2 <- co2 %>%
  rename_with(~ str_remove(., "^X"), starts_with("X"))
co2

co2 <- co2 %>%
  mutate(across(`2003`:`2006`, as.numeric))

# Let's start with removing the unnecessary year columns.

co2_cleaned <- co2 %>%
  select(-any_of(as.character(1800:1951)), -any_of(as.character(2008:2100)))
head(co2_cleaned)

# Excellent, now let's flip the data to long formatting

long_co2 <- co2_cleaned %>%
  pivot_longer(
    cols = -country,            # Select all columns to pivot
    names_to = "Year",              # Name for the new column containing years
    values_to = "Co2"               # Name for the new column containing values
  ) %>%
  mutate(year = as.numeric(Year))
long_co2

# Now let's filter the data to be in the same 5 year increment

filtered_co2 <- long_co2 %>%
  filter(year %in% seq(1952, 2007, by = 5))

# Then finally, merge co2 into the same dataset
merged_data <- gapminder %>%
  left_join(filtered_co2, by = c("country", "year"))
merged_data





```

• Create one or multiple animation(s) to test your hypothesis. Label selected nodes, relevant to your hypothesis, in your animated scatterplots. (3 points)

```{r}
# Let's create our basic plots and store them into variables

p <- ggplot(merged_data,
       aes(x = gdpPercap, y = Co2, size = pop, color = continent)) +
      geom_point() +
      scale_x_log10() +
      scale_color_viridis_d() +
      labs(x = "GDP per Capitta", y = "Co2")
p



p2<- p + facet_wrap(~continent)

p2 + transition_time(year) + labs(title = "Year:{frame_time}")

```

• State your observations regarding your hypothesis. (1 points)

```{r}
# I decided to add size based on population and add a facet_wrap to my final model to see if there were any continent specific issues.

# My final result is that there is a general trend towards higher CO2 emissions with higher GDP per capita and a general trend that my hypothesis was correct. However, there seems to be continent specific problems. Continents with a higher amount of lower GDP countries increased CO2 emissions significantly slower than continents with a lower amount of low gdp countries (i.e. the differences between Africa and Europe.)

```
